WSClean: an implementation of a fast, generic wide-field imager

### 2014-02-04 articles/2014/Article-WSClean/wsclean.tex
Article
Work on article

@@

\begin{figure*}%
\begin{subfigure}{.5\linewidth}%
\includegraphics[width=\linewidth]{img/benchmark-nsamples/nsamples}%
\caption{}\label{fig:timing-nsamples}%
\end{subfigure}%
\begin{subfigure}{.5\linewidth}%
\includegraphics[width=\linewidth]{img/benchmark-zenith-angle/za}%
\caption{}\label{fig:timing-zenith-angle}%
\end{subfigure}\\
\begin{subfigure}{.3\linewidth}%
\includegraphics[width=\linewidth]{img/benchmark-resolution/resolution}
\caption{}
\label{fig:timing-resolution}
\end{subfigure}%
\begin{subfigure}{.3\linewidth}%
\includegraphics[width=\linewidth]{img/benchmark-channels/channels}
\caption{}
\label{fig:timing-channels}
\end{subfigure}%
\begin{subfigure}{.3\linewidth}%
\includegraphics[width=\linewidth]{img/benchmark-fov/fov}
\caption{}
\label{fig:timing-fov}
\end{subfigure}%
\caption{Imaging performance as a function of several configuration parameters. Error bars show 5$\sigma$ level. }\label{fig:timings}
\end{figure*}

@@

\subsection{Derivation of computing cost formulae}
Based on the expected cost terms described in Sec.~\ref{sec:time-complexity}, we fit analytical functions to the CASA and WSClean measurements. Several functions with different free parameters were tested, and formulae with minimum amount of parameters are selected that still follow the trend of the measurements and have reasonably small errors. The measurements are weighted with the inverse standard deviation. The single outlying measurement of CASA in Fig.~\ref{fig:timing-nsamples} is removed.

For WSClean's time cost $t_\textrm{WSClean}$ we find 
\begin{align}\notag
t_\textrm{WSClean}& (w_e,N_\textrm{vis},N_\textrm{freq},N_\textrm{pix}) = \\ \notag
& 0.000326 N_\textrm{pix}^2 \log N_\textrm{pix} (w_e + 0.000212) +\\
& 0.252 N_\textrm{vis} - 7.64
\end{align}
and for CASA we find
\begin{equation}
t_\textrm{CASA}(w_e,N_\textrm{vis},N_\textrm{freq},N_\textrm{pix}) = .
\end{equation}
Parameter $w_e$ is the maximum $w$ value and is esimated with
\begin{equation}
 w_e = \frac{\max |\mathbf{x}| \sin za + \max z \cos za}{\lambda \max |\mathbf{x}|} \left(1.0 - \cos(\frac{1}{2}\textrm{FOV})\right),
\end{equation}
where $\max |\mathbf{x}|$ is the maximum baseline length, $\max z$ is the maximum height difference between antennas and $\lambda$ is the wavelength.

The prolate spheroidal wave function (PSWF) is generally considered to be the optimal windowing function for gridding \citep{fourier-kernel-selection-1991}. CASA convolves samples with a PSWF of 7 pixels total width during gridding. When using a variable kernel size, a PSWF is quite complicated and computational expensive to calculate. WSClean currently uses a Kaiser-Bessel (KB) window function, which is easy and fast to compute, and is a good approximation for the PSWF \citep{fourier-kernel-selection-1991}.

\begin{figure}
\begin{center}
\includegraphics[width=8cm]{img/benchmark-kernelsize/kernel}
\caption{...}
\label{fig:timing-kernelsize}
\end{center}
\end{figure}

The type of window function has no effect on the gridding performance, but gridding has a quadratic time dependency on the size of the kernel. Fig.~\ref{fig:timing-kernelsize} shows that this quadratic dependency becomes apparent for kernel sizes $\gtrapprox 15$. Decreasing the kernel size to values below $7$ has little effect on performance, because with small kernels reading the samples takes more time than gridding. We have not noticed much benefit of larger kernels except in rare cases where a bright source lies just outside the imaged field of view. Therefore, WSClean uses a default of 7 pixels for the gridding kernel. It can be increased when necessary.

@@

\section*{Acknowledgments}
...

>>>>

The prolate spheroidal wave function (PSWF) is generally considered to be the optimal windowing function for gridding \citep{fourier-kernel-selection-1991}. CASA convolves samples with a PSWF of 7 pixels total width during gridding. When using a variable kernel size, a PSWF is quite complicated and computational expensive to calculate. WSClean currently uses a Kaiser-Bessel (KB) window function, which is easy and fast to compute, and is a good approximation for the PSWF \citep{fourier-kernel-selection-1991}. After some benchmarks (see Appendix~\ref{sec:gridding-appendix}) we found that 7 pixels is a good compromise between performance and accuracy, and selected this as the default for WSClean.

@@

\begin{figure*}%
\begin{subfigure}{.36\linewidth}%
\includegraphics[width=\linewidth]{img/benchmark-nsamples/nsamples}%
\caption{}\label{fig:timing-nsamples}%
\end{subfigure}%
\begin{subfigure}{.36\linewidth}%
\includegraphics[width=\linewidth]{img/benchmark-zenith-angle/za}%
\caption{}\label{fig:timing-zenith-angle}%
\end{subfigure}\\
\begin{subfigure}{.36\linewidth}%
\includegraphics[width=\linewidth]{img/benchmark-resolution/resolution}
\caption{}\label{fig:timing-resolution}%
\end{subfigure}%
\hspace{-.05\linewidth}\begin{subfigure}{.36\linewidth}%
\includegraphics[width=\linewidth]{img/benchmark-channels/channels}
\caption{}\label{fig:timing-channels}%
\end{subfigure}%
\hspace{-.05\linewidth}\begin{subfigure}{.36\linewidth}%
\includegraphics[width=\linewidth]{img/benchmark-fov/fov}
\caption{}
\label{fig:timing-fov}
\end{subfigure}%
\caption{Imaging performance as a function of several configuration parameters. Error bars show 5$\sigma$ level. }\label{fig:timings}
\end{figure*}

@@

\subsection{Derivation of computing cost formulae}
Based on the expected cost terms described in Sec.~\ref{sec:time-complexity}, we derive analytical functions for the CASA and WSClean measurements using least-squares fitting. Several functions with different free parameters were tested, and formulae with minimum amount of parameters are selected that still follow the trend of the measurements and have reasonably small errors. The measurements are weighted with the inverse standard deviation. The single outlying measurement of CASA in Fig.~\ref{fig:timing-nsamples} is removed.

For WSClean's time cost $t_\textrm{WSClean}$ we find 
\begin{align}\notag
t_\textrm{WSClean}& (w_e,N_\textrm{vis},N_\textrm{freq},N_\textrm{pix}) = \\ \notag
& 0.000326 N_\textrm{pix}^2 \log N_\textrm{pix} (w_e + 0.000212) +\\
& 0.252 N_\textrm{vis} - 7.64
\end{align}
and for CASA we find
\begin{equation}
t_\textrm{CASA}(w_e,N_\textrm{vis},N_\textrm{freq},N_\textrm{pix}) = .
\end{equation}
Parameter $w_e$ is the maximum $w$ value and is estimated with
\begin{equation}
 w_e = \frac{\max |\mathbf{x}| \sin za + \max z \cos za}{\lambda \max |\mathbf{x}|} \left(1.0 - \cos(\frac{1}{2}\textrm{FOV})\right),
\end{equation}
where $\max |\mathbf{x}|$ is the maximum baseline length, $\max z$ is the maximum height difference between antennas and $\lambda$ is the wavelength.

@@

\section*{Acknowledgments}
...
\appendix

\section{Optimal gridding kernel size} \label{sec:gridding-appendix}
\begin{figure}
\begin{center}
\includegraphics[width=8cm]{img/benchmark-kernelsize/kernel}
\caption{...}
\label{fig:timing-kernelsize}
\end{center}
\end{figure}
As discussed in \S\ref{sec:gridding}, WSClean uses a Kaiser-Bessel window against aliasing. The type of window function has no effect on the gridding performance, but the size of the kernel affects gridding performance quadratically. Fig.~\ref{fig:timing-kernelsize} shows that this quadratic relation becomes significant for kernel sizes $\gtrapprox 10$. Decreasing the kernel size to values below $7$ has little effect on performance, because for small kernels the visibility-reading rate is lower than the gridding rate. We have not noticed much benefit of larger kernels except in rare cases where a bright source lies just outside the imaged field of view. Therefore, WSClean uses a default of 7 pixels for the gridding kernel. It can be increased when necessary. Different hardware might have a slightly different optimum because of different reading and calculation performance.
